Cultural Transmission Can Explain the Evolution of Cooperative Behavior

### 2020-07-16 ms/main.tex
changes and additions by MWF

% Introduction
\section*{Introduction}
Cooperative behavior can harm the individual's fitness and increase the fitness of its competitor~\citep{axelrod1981evolution}.
Yet, cooperative behavior occurs in many non-human animals~\citep{dugatkin1997cooperation}, for example rats~\citep{rice1962altruism} and birds~\citep{krams2008experimental}.
Therefore, the evolution of cooperative behavior is an important open question in evolutionary biology.
\\
Cultural evolution is an evolutionary theory of social change.
Culture has significant impact on the behavior of humans~\citep{ihara2004cultural,jeong2018bronze} as well as non-human animals~\citep{bonner2018evolution}.
Under the view of cultural evolution, an individual can acquire its behavior from another individual in its social group through learning or other modes of cultural transmission \citep{richerson2008not}.
Here we attempt to determine to what extent cultural transmission can explain the evolution of cooperative behavior.

@@

\paragraph{Kin selection theory} suggests that natural selection can favor cooperative behavior between kin. The importance of relatedness to the evolution of cooperation and altruism was shown by~\citet{hamilton1964genetical}. According to Hamilton, kin selection causes allele to increase in frequency when the the reproductive cost to the actor, $c$, is less than the benefit to the recipient, $b$, times the genetic relatedness between the recipient and the actor, $r$.
This is also known as Hamilton's Rule:

@@

There is an ongoing debate about to what extent kin selection explains evolution of cooperation and altruism.
It has been suggested that kin selection to explain the cooperative behavior of eusocial insects like the honey bee.
The most significant argument against kin selection is that cooperation can evolve with zero relatedness~\citep{wilson2005kin}. This makes Hamilton's rule incomplete according to \citet{wilson2005kin}. \citet{foster2006kin} reject this claim. They argue that altruism without relatedness can not evolve. They refer us to Hamilton who claimed that relatedness can arise without recent common ancestry. 
\citet{wilson2005kin} also criticizes kin selection on the grounds that environmental or ecological factors probably be more important than relatedness in determining social actions. On the other hand, \citet{foster2006kin} argue that kin selection does not ignore ecology. Hamilton's rule shows
that environmental factors causing a high benefit-to-cost ratio will favor cooperation.

\paragraph{Reciprocity} suggests repeating interactions or individual recognition as key factors for explaining the evolution of cooperation. In \emph{direct reciprocity} there are a repeated encounters between the same two individuals. In every encounter, each player has a choice between cooperation and defection. If I cooperate now, you may cooperate later. Hence, it may pay off to cooperate.
This game-theoretic framework is known as the repeated Prisoner's Dilemma. 
%A well known strategy to play this game is called  'tit-for-tat'. This strategy always starts with a cooperation, then it does whatever the other player has done in the previous round: a cooperation for a cooperation, a defection for a defection. There a unlimited number of possible strategies to play this game. However, 
Direct reciprocity can only lead to the evolution of cooperation if the cost is less than the benefit $b$ times the probability for another encounter between the same two individuals, $w$,

@@

Direct reciprocity assumes that both players are in a position to cooperate. Direct reciprocity can not explain cooperation in asymmetric interactions. In humans, such interactions happen often, for example humans often donate money.

\paragraph{Indirect reciprocity} has been suggested to explain this behavior.
\citet{nowak2006five} claims that direct reciprocity is like a barter economy based on the immediate exchange of goods, while indirect reciprocity resembles the invention of money. The money that "fuels the engines" of indirect reciprocity is reputation. 
However, Reciprocity assume repeating interactions and therefore, has difficulty explaining evolution of cooperation if the no repeating interactions occurs.

\paragraph{Group selection theory} posits that cooperation is favored because of the advantage to the whole group, if selection acts at the group level in addition to the individual level. A common model for group selection work as is: the population is divided into groups. In each groups there are cooperators, which help to other group members and defectors which do not help.  % TODO add reference
Individuals reproduce proportional to their fitness. Offspring are added to the same group.
If a group reaches a certain size it can split to two groups. A group that grow faster will split more often.
Groups of cooperators are growing faster than group of defectors.
Therefore, cooperation can evolve in this model when the cost $c$ is less than the benefit $b$ times the ratio between the  the number of groups $m$ and the sum of $m$ and the maximum group size $n$,

@@

Group selection was criticized by biologists advocate gene-centered view of evolution. Group selection has been criticized due to the fact that the trait like cooperation evolves in the total population. According to natural selection, if cooperation took over the population it must have better fitness. However, in group selection the fitness of cooperator in the individual level is lower. The fact that a trait with a lower fitness took over the population is a contradiction. \citet{eldakar2011eight} reject this argument. They believe that this argument is a tautology and does not qualify as an argument against group selection. The distinction between individual and group selection requires a comparison of fitness differentials within and among groups in a multi group population. When a trait is evolve by group selection, despite the fact that it has lower fitness within the group, it has a better fit, all thing considered.

\paragraph{}
All the above theories assume that cooperation is genetically determined. This raises the question, is it possible that cooperation is determined by non-genetic factors?
Recent work by \citet{lewin2017microbes} sheds some light on this question. 
\citet{lewin2017microbes} have hypothesised that microbes that manipulate their hosts to act altruistically can be favored by selection, and may play a role in the widespread occurrence of cooperative behavior. Indeed, it has been shown that microbes can mediate behavioral changes in their hosts~\citep{poulin2010parasite,dobson1988population}. Therefore, natural selection on microbes may favor manipulation of the host so that it cooperates with others. Microbes can be transmitted \emph{horizontally} from one host to another during host interactions. Following horizontal transfer, the recipient host may carry microbes that are closely related to the microbes of the donor host, even when the two hosts are (genetically) unrelated~\citep{lewin2017microbes}. Microbes can also transfer vertically, from parent to offspring. % yr: cite 10.1126/science.aat7164
As a result, a microbe that induces its host to cooperate with another host and thereby increases the other host fitness will  increase the vertical transmission of the microbes of the receiving individual. Kin selection among microbes could therefore favor microbes that induce cooperative behavior in their hosts, thereby increasing the transmission of their microbial kin.
%\\While the hypothesis that microbes could manipulate their hosts to act altruistically is mathematically feasible~\citep{lewin2017microbes}, there are still no empirical evidence of whether microbes indeed mediate altruistic behavior of their hosts.

% subsection
\subsection*{Cultural evolution of cooperation}
\citet{lewin2017microbes} have demonstrated that \emph{non-vertical transmission} can help to explain the evolution of cooperative behavior. 
Non-vertical transmission could be either a horizontal or oblique. Horizontal transmission occurs between individuals from the same generation. Oblique transmission occurs from an adult to an offspring. 
Evolution under either of these transmission models is can be be more rapid than under pure vertical transmission~\citep{ram2018evolution}. % (yr: not sure that this is a good reference)

Here we suggest a model in which behavioral changes are mediated by cultural transmission that can occur during social interaction. For example, if an individual interacts with a cooperative individual, it might learn that cooperation is a positive behavior and will cooperate in the future. Surprisingly, some of the analysis made by \citet{lewin2017microbes} can be applied to cultural transmission, because cultural transmission is mathematically akin to transmission of infectious diseases~\citep{cavalli1981cultural}.

Here, we hypothesize that non-vertical cultural transmission can explain the evolution of cooperation. We are develop cultural evolution models that include both vertical and non-vertical transmission of cooperation and investigate these models using both mathematical analysis and simulations.

@@

Individuals interact in a social interaction modeled as a prisoner's dilemma.
Specifically, individuals interact in pairs, cooperators pay a fitness cost $0<c<1$, and their partner gains a fitness benefit $b$, where we assume $b>c$. The following payoff matrix shows the fitness of an individual with phenotype $\phi_1$ when interacting with a partner with phenotype $\phi_2$ ($b>c>0$):

@@

Social interactions occur randomly.
So, two individuals with phenotype $A$ interact with probability $\hat{p}^2$, two individuals with phenotype $B$ interact with probability $(1-\hat{p})^2$, and two individuals with different phenotypes interact with probability $2\hat{p}(1-\hat{p})$.

@@

Horizontal cultural transmission occurs between peers. 
It may occur between social partners with probability $\alpha$, or between a random pair with probability $1-\alpha$ (see~\textbf{\autoref{fig:horizontal}}).
Horizontal transmission is not always successful, as one peer may reject the other's phenotype. The probability for successful transmission of phenotypes $A$ and $B$ are $T_A$ and $T_B$, respectively.\\
\autoref{table:1} contains the probability of first interactor to be A following interaction by interaction type.
Therefore, the frequency $p'$ of phenotype $A$ among adults in the next generation, after horizontal transmission, is 
\begin{equation}\label{eq:nextgen_adults}
\begin{aligned}
p'
& = \hat{p}^2 [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A] \;,
\end{aligned}
\end{equation}
which can be simplified into
\begin{equation}\label{eq:nextgen_adults_slimpify}
p' = \hat{p}^2(T_B-T_A) + \hat{p}(1+T_A-T_B) .
\end{equation}

The frequency of $A$ among parents follows a similar dynamic, but also includes the effect of natural selection, and is therefore
\begin{equation}\label{eq:nextgen_parents}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) (1-c) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} (1+b) [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A] \;,
\end{aligned}
\end{equation}
where fitness values are taken from \autoref{table:1}, and 
\begin{equation} \label{eq:mean_fitness}
\bar{w} = \hat{p} w_A + (1-\hat{p}) w_B = 1 + \hat{p}(b-c)
\end{equation}
is the population mean fitness.
Eq~\ref{eq:nextgen_parents} can be slimpified to
\begin{equation}\label{eq:nextgen_parents_simplified}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) \big(1-(1-\hat{p})(1-\alpha)T_B)\big) \\
& + \hat{p}(1-\hat{p}) (1-c) \big(\hat{p}(1-\alpha)T_B+1-T_B\big) \\
& + (1-\hat{p})\hat{p} (1+b) \big(\hat{p}(1-\alpha) + \alpha\big) T_A \\
& + (1-\hat{p})^2 \hat{p} (1-\alpha) T_A \;.
\end{aligned}
\end{equation}

\begin{table}[H]
\begin{tabular}{|c|c|c|c|c|}
\hline
\textbf{Interaction Type} & \textbf{Frequency}    & \textbf{Fitness}      & \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}Probability of first interactor\\  to be A following interaction\end{tabular}}}                                                                                                     \\ \hline
\multicolumn{1}{|l|}{}    & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \textbf{\begin{tabular}[c]{@{}c@{}}Horizontal transmission\\  from partner, \\ probability $\alpha$\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Horizontal transmission \\ from population, \\ probability $1-\alpha$\end{tabular}} \\ \hline
$AxA$                     & $\hat{p}^2$           & $1+b-c$               & $1$                                                                                                               & $\hat{p}+(1-\hat{p})(1-T_B)$                                                                                           \\ \hline
$AxB$                     & $\hat{p}(1-\hat{p})$  & $1-c$                 & $1-T_B$                                                                                                           & $\hat{p}+(1-\hat{p})(1-T_B)$                                                                                           \\ \hline
$BxA$                     & $\hat{p}(1-\hat{p})$  & $1+b$                 & $T_A$                                                                                                             & $\hat{p}T_A$                                                                                                           \\ \hline
$BxB$                     & $(1-\hat{p})^2$       & $1$                   & $0$                                                                                                               & $\hat{p}T_A$                                                                                                           \\ \hline
\end{tabular}
\caption{Fitness and probabilityof first interactor to be A following interaction by interaction type}
\label{table:1}
\end{table}

\begin{figure}[b]
  \centering
  \includegraphics[scale=1]{figure1.pdf}
  \caption{\textbf{Cultural horizontal transmission.} Transmission occurs between interaction partners with probability $\alpha$ (left) or between two random peers with probability $1-\alpha$.}
  \label{fig:horizontal}
\end{figure}

@@

We start by inspecting specific cases, for which we can derive general results. 
Afterwards, we use numerical simulation to analyze more complex cases.

@@

\textbf{Result 1: Vertical and horizontal transmission of cooperation.}
If 
\begin{equation} \label{eq:unequal_transmission}
c \cdot (1-T_B) < b \cdot  \alpha T_A  + (T_A - T_B) [1 + b\tilde{p}(1-\alpha)] ,
\end{equation}
then $\tilde{p}' > \tilde{p}$, and the frequency of the cooperator phenotype $A$ among parents increases every generation.

@@

Therefore, because all the parameters are all positive, cooperation cannot take over the population (and furthermore will become extinct) if cooperators doesn't have a horizontal transmission bias, i.e. if $T_A\leq T_B$.
When such a bias does exist, $T_A > T_B$, then cooperation will evolve if $\tilde{p} > \tilde{p}^{*}$, where (see also ~\textbf{\autoref{fig:results_b}})
\begin{equation} 
\begin{split} \label{eq:horizontal_transmission_from_population_equilibrium}
\tilde{p}^{*} & = \frac{c}{b} \cdot \frac{1-T_B}{T_A-T_B} - \frac{1}{b}.
\end{split}
\end{equation}
A sufficient condition for evolution of cooperation is that $\tilde{p}^{*} < 0$, which occurs if
\begin{equation} \label{eq:sufficient_horizontal_transmission_from_population}
c < \frac{T_A-T_B}{1-T_B} .  
\end{equation}

@@

With only oblique and horizontal transmission, i.e. $v = 0$, eq.~\ref{eq:horizontal} becomes $\hat{p}=p$ and eq.~\ref{eq:nextgen_adults_slimpify} becomes % TODO check this, I changed the ref from nextgen_parents_simplified to nextgen_adults_slimpify
\begin{equation}  \label{eq:nextgen_parents_oblique_only}
p' = p^2 (T_B-T_A) + p (1+T_A-T_B) .
\end{equation}

We find the following results.

\textbf{Result 2: Oblique and horizontal transmission of cooperation.} 
If there is a horizontal transmission bias in favor of cooperation,
\begin{equation} \label{eq:oblique_only_result}
T_A > T_B, 
\end{equation}
then $p'>p$, and the frequency of the cooperator phenotype among adults increases every generation.
Therefore, cooperation will evolve if the cooperator phenotype has a horizontal transmission bias (see~\textbf{\autoref{fig:results_c}}).

@@

In this case $0<v<1$, and the equation system is more complex.
Therefore, we focus on local, rather than global, stability.\\
To proceed, we note that 
eq.~\ref{eq:horizontal} can give $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$,
eq.~\ref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\hat{p}$, and 
eq.~\ref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$.
We then determine the equilibria of this equation, that is, solutions for $\hat{p}' = \hat{p}$, and analyse their local stability: an equilibrium $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})$ at the equilibrium is negative, $f'(\hat{p}^*)<0$. %TODO show this is necessary and sufficient

We start with the simpler case of symmetric horizontal transmission, $T=T_A=T_B$. 
We apply eqs.~\ref{eq:horizontal}, \ref{eq:nextgen_adults_slimpify}, and \ref{eq:nextgen_parents_simplified} to define 
\begin{equation} \label{eq:equal_horizontal_transmission}
\begin{aligned}
  %\hat{p}' - \hat{p} = 
  %\hat{p} ^2(-\alpha bvT + cv(1-T)) + \hat{p} (\alpha bvT -cv(1-T)) = 
  f(\hat{p}) &= 
  \bar{w}(\hat{p}' - \hat{p}) = \\
  &\hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big]. % shorter version
\end{aligned}
\end{equation}

The equilibria are solutions to $f(\hat{p})=0$, or $\hat{p}' = \hat{p}$.
It is easy to verify that fixation of either phenotype, $\hat{p} =  0$ and $\hat{p} = 1$, is an equilibrium.
Since the derivative of $f'(\hat{p}$ is 
\begin{equation}
f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big],
\end{equation}
then the condition for local stability of $\hat{p}=1$ is
\begin{equation} \label{eq:derivative_of_phattag-phat}
	f'(1) =
% 2\cdot\hat{p} (-\alpha bvT + cv(1-T)) + (\alpha bvT -cv(1-T)).
%	(1- 2\hat{p}) (\alpha bvT - cv(1-T)) \mid_{\hat{p}=1} = 
	-\alpha bvT + cv(1-T) < 0,
\end{equation}
which gives us the following result.

>>>>

% Introduction
\section*{Introduction}
Cooperative behavior can harm an individual's fitness and increase the fitness of its conspecifics or competitors~\citep{axelrod1981evolution}.
Nevertheless, cooperative behavior appears to occur in many non-human animals~\citep{dugatkin1997cooperation}, for example rats~\citep{rice1962altruism} and birds~\citep{krams2008experimental}.
Evolution of cooperative behavior remains  an important  conundrum in evolutionary biology.
\\
Cultural evolution is an evolutionary theory of individual and group-level change in beliefs, behaviors, and norms \citep{cavalli1981cultural}.
Culture has significant impact on the behavior of humans~\citep{ihara2004cultural,jeong2018bronze} as well as non-human animals~\citep{bonner2018evolution}.
Cultural transmission allows an individual to acquire attitudes and behavioral traits from other individuals in its social group through imitation, learning, or other modes of communication \citep{cavalli1981cultural,richerson2008not}.
Here we attempt to determine to what extent cultural transmission can explain the evolution of cooperative behavior.

@@

\paragraph{Kin selection theory} posits that natural selection can favor cooperation between related individuals. The importance of relatedness to the evolution of cooperation and altruism was shown by~\citet{hamilton1964genetical}. According to Hamilton,  an allele that determines altruistic behavior should increase in frequency when  the reproductive cost to the actor that cooperates, $c$, is less than the benefit to the recipient, $b$, times the genetic relatedness between the recipient and the actor, $r$.
This is also known as Hamilton's Rule:

@@

There is an ongoing debate about to what extent kin selection explains evolution of cooperation and altruism.
It has been suggested that kin selection can explain the cooperative behavior of worker castes of eusocial insects like the honey bee.
The most significant argument against kin selection is that in some cases cooperation among unrelated individuals appears to have evolved~\citep{wilson2005kin}. This makes Hamilton's rule incomplete according to \citet{wilson2005kin}, although \citet{foster2006kin} reject this claim and argue that altruism cannot evolve without relatedness. They refer  to Hamilton, who claimed that relatedness can arise without recent common ancestry. %MWF: How?
\citet{wilson2005kin} also criticizes kin selection on the grounds that environmental or ecological factors are probably more important than relatedness in determining social actions. Although \citet{foster2006kin} argue that kin selection does not ignore ecology. Hamilton's rule suggests
that environmental factors causing a high benefit-to-cost ratio will favor cooperation.

\paragraph{Reciprocity} entails that repeated interactions or individual recognition are key components of the evolution of cooperation. In \emph{direct reciprocity} there are repeated encounters between the same two individuals, where at every encounter, each player has a choice between cooperation and defection: if I cooperate now, you may cooperate later. Hence, it may pay off to cooperate.
This game-theoretic framework, known as the repeated Prisoner's Dilemma, 
%A well known strategy to play this game is called  'tit-for-tat'. This strategy always starts with a cooperation, then it does whatever the other player has done in the previous round: a cooperation for a cooperation, a defection for a defection. There a unlimited number of possible strategies to play this game. However, 
can only lead to the evolution of cooperation if the cost is less than the benefit $b$ times the probability of another encounter between the same two individuals, $w$, 
%MWF: Referencess?

@@

Direct reciprocity assumes that both players are in a position to cooperate, but it can not explain cooperation in asymmetric interactions such as human philanthropy.

\paragraph{Indirect reciprocity} has also been suggested as an explanation of this behavior.
\citet{nowak2006five} claims that direct reciprocity is like a barter economy based on the immediate exchange of goods, while indirect reciprocity resembles the invention of money. 
%MWF: More refs
The money that ``fuels the engines'' of indirect reciprocity is reputation. 
However, reciprocity assumes repeated interactions and therefore has difficulty in explaining the evolution of cooperation if  interactions are not repeated.

\paragraph{Group selection theory} posits that cooperation is favored because it imparts an advantage to the whole group, if selection acts at the group level in addition to the individual level. A common model for group selection divides the population into groups in which there are cooperators that help  other group members and defectors that do not.  % TODO add reference
Individuals reproduce proportionally to their fitness, and offspring are added to the same group as their parents.
If a group reaches a certain size it can split to two groups, so groups that grow faster will split more often.
Groups with cooperators  grow faster than groups without cooperators, and
cooperation can evolve in this model when the cost $c$ is less than the benefit $b$ times the ratio between the  the number of groups $m$ and the sum of $m$ and the maximum group size $n$,

@@

Group selection has been criticized by biologists who advocate a gene-centered view of evolution. Group selection has also been criticized because a trait such as cooperation evolves in the total population. For cooperation to take over the population it must have higher fitness, while under group selection the fitness of cooperators at the individual level is lower. Thus a trait with a lower fitness taking over the population is a contradiction. \citet{eldakar2011eight} reject this argument claiming that it is a tautology and does not qualify as an argument against group selection. The distinction between individual and group selection requires a comparison of fitness differentials within and between groups in a multi-group population, and when a trait  evolves by group selection, despite having lower fitness within a group, that group might have higher average fitness in competition with other groups, all things considered.

\paragraph{}
All the above theories assume that cooperation is genetically determined, which raises the question, is it possible that cooperation is determined by non-genetic factors? \citet{feldman1985gene} introduced the first model for the evolution of altruism by cultural transmission. They showed that under vertical (parent-to-offspring) cultural transmission, Hamilton's rule does not govern the evolution of parent-to-offspring or sib-to-sib altruism.
Recent work by \citet{lewin2017microbes} also sheds some light on this question. 
They hypothesise that microbes that manipulate their hosts to act altruistically can be favored by selection, and may play a role in the widespread occurrence of cooperative behavior. Indeed, it has been shown that microbes can mediate behavioral changes in their hosts~\citep{poulin2010parasite,dobson1988population}. Therefore, natural selection on microbes may favor manipulation of the host so that it cooperates with others. Microbes can be transmitted \emph{horizontally} from one host to another during host interactions, and following horizontal transfer, the recipient host may carry microbes that are closely related to the microbes of the donor host, even when the two hosts are (genetically) unrelated~\citep{lewin2017microbes}. Microbes can also be transferred vertically, from parent to offspring, and % yr: cite 10.1126/science.aat7164
a microbe that induces its host to cooperate with another host and thereby increases the latter's fitness will  increase the vertical transmission of the microbes of the receiving individual. Kin selection among microbes could therefore favor microbes that induce cooperative behavior in their hosts, thereby increasing the transmission of their microbial kin.
%\\While the hypothesis that microbes could manipulate their hosts to act altruistically is mathematically feasible~\citep{lewin2017microbes}, there are still no empirical evidence of whether microbes indeed mediate altruistic behavior of their hosts.

% subsection
\subsection*{Cultural evolution of cooperation}
\citet{lewin2017microbes} have demonstrated that \emph{non-vertical transmission} can help to explain the evolution of cooperative behavior. 
Non-vertical transmission may be either a horizontal or oblique: horizontal transmission occurs between individuals from the same generation, while oblique transmission occurs from an adult to an offspring. 
Evolution under either of these transmission models  can be be more rapid than under pure vertical transmission~\citep{cavalli1981cultural,ram2018evolution}. % (yr: not sure that this is a good reference)

Here we suggest a model in which behavioral changes are mediated by cultural transmission that can occur during social interaction. For example, if an individual interacts with a cooperative individual, it might learn that cooperation is a positive behavior and will cooperate in the future. Some of the analysis made by \citet{lewin2017microbes} can be applied to cultural transmission, because models of cultural transmission are mathematically similar to those for transmission of infectious diseases~\citep{cavalli1981cultural}.

Here, we develop cultural evolution models that include both vertical and non-vertical transmission of cooperation and investigate these models using both mathematical analysis and simulations.  We hypothesize that non-vertical cultural transmission can explain the evolution of cooperation.

@@

Individuals interact according to a prisoner's dilemma.
Specifically, individuals interact in pairs; a cooperator suffers a fitness cost $0<c<1$, and its partner gains a fitness benefit $b$, where we assume $b>c$. The following payoff matrix shows the fitness of an individual with phenotype $\phi_1$ when interacting with a partner of phenotype $\phi_2$ ($b>c>0$):

@@

Social interactions occur randomly:
two individuals with phenotype $A$ interact with probability $\hat{p}^2$, two individuals with phenotype $B$ interact with probability $(1-\hat{p})^2$, and two individuals with different phenotypes interact with probability $2\hat{p}(1-\hat{p})$.

@@

Horizontal cultural transmission occurs between peers. 
It may occur between social partners with probability $\alpha$, or between a random pair with probability $1-\alpha$ (see~\textbf{\autoref{fig:horizontal}}).
Horizontal transmission is not always successful, as one peer may reject the other's phenotype. The probability for successful transmission of phenotypes $A$ and $B$ are $T_A$ and $T_B$, respectively.\\
\autoref{table:1} contains the probability of first interactor to be A following interaction by interaction type.
Therefore, the frequency $p'$ of phenotype $A$ among adults in the next generation, after horizontal transmission, is 
\begin{equation}\label{eq:nextgen_adults}
\begin{aligned}
p'
& = \hat{p}^2 [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A] \;,
\end{aligned}
\end{equation}
which simplifies to
\begin{equation}\label{eq:nextgen_adults_slimpify}
p' = \hat{p}^2(T_B-T_A) + \hat{p}(1+T_A-T_B) .
\end{equation}

The frequency of $A$ among parents (after selection) follows a similar dynamic, but also includes the effect of natural selection, and is therefore
\begin{equation}\label{eq:nextgen_parents}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) (1-c) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} (1+b) [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A] \;,
\end{aligned}
\end{equation}
where fitness values are taken from \autoref{table:1}, and 
\begin{equation} \label{eq:mean_fitness}
\bar{w} = \hat{p} w_A + (1-\hat{p}) w_B = 1 + \hat{p}(b-c)
\end{equation}  %MWF: Define w_A, w_B.
is the population mean fitness.
Eq~\ref{eq:nextgen_parents} can be simplified to
\begin{equation}\label{eq:nextgen_parents_simplified}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) \big(1-(1-\hat{p})(1-\alpha)T_B)\big) \\
& + \hat{p}(1-\hat{p}) (1-c) \big(\hat{p}(1-\alpha)T_B+1-T_B\big) \\
& + (1-\hat{p})\hat{p} (1+b) \big(\hat{p}(1-\alpha) + \alpha\big) T_A \\
& + (1-\hat{p})^2 \hat{p} (1-\alpha) T_A \;.
\end{aligned}
\end{equation}

\begin{table}[H]
\begin{tabular}{|c|c|c|c|c|}
\hline
\textbf{Interaction Type} & \textbf{Frequency}    & \textbf{Fitness}      & \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}Probability of first interactor\\  to be A %MWF: What is A?
 following interaction\end{tabular}}}                                                                                                     \\ \hline
\multicolumn{1}{|l|}{}    & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \textbf{\begin{tabular}[c]{@{}c@{}}Horizontal transmission\\  from partner, \\ probability $\alpha$\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Horizontal transmission \\ from population, \\ probability $1-\alpha$\end{tabular}} \\ \hline
$AxA$                     & $\hat{p}^2$           & $1+b-c$               & $1$                                                                                                               & $\hat{p}+(1-\hat{p})(1-T_B)$                                                                                           \\ \hline
$AxB$                     & $\hat{p}(1-\hat{p})$  & $1-c$                 & $1-T_B$                                                                                                           & $\hat{p}+(1-\hat{p})(1-T_B)$                                                                                           \\ \hline
$BxA$                     & $\hat{p}(1-\hat{p})$  & $1+b$                 & $T_A$                                                                                                             & $\hat{p}T_A$                                                                                                           \\ \hline
$BxB$                     & $(1-\hat{p})^2$       & $1$                   & $0$                                                                                                               & $\hat{p}T_A$                                                                                                           \\ \hline
\end{tabular}
\caption{Fitness and probability of the first interactor to be A following interaction by interaction type} %MWF: What is A?
\label{table:1}
\end{table}

\begin{figure}[b]
  \centering
  \includegraphics[scale=1]{figure1.pdf}
  \caption{\textbf{Cultural horizontal transmission.} Transmission occurs between interacting partners with probability $\alpha$ (left) or between two random peers with probability $1-\alpha$.}
  \label{fig:horizontal}
\end{figure}

@@

We start by studying specific cases, for which we can derive general results, and then 
we use numerical simulation to analyze more complex cases.

@@

\textbf{Result 1: Vertical and horizontal transmission of cooperation.}
From eq.~\ref{eq:nextgen_parents_vertical_only}, if
\begin{equation} \label{eq:unequal_transmission}
c \cdot (1-T_B) < b \cdot  \alpha T_A  + (T_A - T_B) [1 + b\tilde{p}(1-\alpha)] ,
\end{equation}
then $\tilde{p}' > \tilde{p}$, and the frequency of the cooperator phenotype $A$ among parents increases every generation.

@@

Therefore, since all the parameters are  positive, cooperation cannot take over the population (and furthermore will become extinct) if cooperators do not have a horizontal transmission bias, i.e. if $T_A\leq T_B$.
When such a bias does exist, $T_A > T_B$, then cooperation will evolve if $\tilde{p} > \tilde{p}^{*}$, where (see also ~\textbf{\autoref{fig:results_b}})
\begin{equation} 
\begin{split} \label{eq:horizontal_transmission_from_population_equilibrium}
\tilde{p}^{*} & = \frac{c}{b} \cdot \frac{1-T_B}{T_A-T_B} - \frac{1}{b}.
\end{split}
\end{equation}
A sufficient condition for evolution of cooperation is that $\tilde{p}^{*} < 0$, namely
\begin{equation} \label{eq:sufficient_horizontal_transmission_from_population}
c < \frac{T_A-T_B}{1-T_B} .  
\end{equation}

@@

With only oblique and horizontal transmission, i.e. $v = 0$, eq.~\ref{eq:horizontal} becomes $\hat{p}=p$ and eq.~\ref{eq:nextgen_adults_slimpify} becomes % TODO check this, I changed the ref from nextgen_parents_simplified to nextgen_adults_slimpify
\begin{equation}  \label{eq:nextgen_parents_oblique_only}
p' = p^2 (T_B-T_A) + p (1+T_A-T_B) ,
\end{equation}

which gives the following result.

\textbf{Result 2: Oblique and horizontal transmission of cooperation.} 
If there is a horizontal transmission bias in favor of cooperation, namely
\begin{equation} \label{eq:oblique_only_result}
T_A > T_B, 
\end{equation}
then $p'>p$, and the frequency of the cooperator phenotype among adults increases every generation.
Therefore, cooperation will evolve if the cooperator phenotype has a horizontal transmission bias (see~\textbf{\autoref{fig:results_c}}).

@@

In this case $0<v<1$, and the recursion system is more complex.
Therefore, we focus on local, rather than global, stability.
To proceed, we note that 
eq.~\ref{eq:horizontal} can give $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$,
eq.~\ref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\hat{p}$, and 
eq.~\ref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$.
We then determine the equilibria, which are solutions of $\hat{p}' = \hat{p}$, and analyse their local stability: an equilibrium $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})$ at the equilibrium is negative, $f'(\hat{p}^*)<0$. %MWF: Define f(\hat p).

We start with the simple case of symmetrical horizontal transmission, $T=T_A=T_B$ and 
%MWF: Write the full recursion.
apply eqs.~\ref{eq:horizontal}, \ref{eq:nextgen_adults_slimpify}, and \ref{eq:nextgen_parents_simplified} to obtain 
\begin{equation} \label{eq:equal_horizontal_transmission}
\begin{aligned}
  %\hat{p}' - \hat{p} = 
  %\hat{p} ^2(-\alpha bvT + cv(1-T)) + \hat{p} (\alpha bvT -cv(1-T)) = 
  f(\hat{p}) &= 
  \bar{w}(\hat{p}' - \hat{p}) = \\
  &\hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big]. % shorter version
\end{aligned}
\end{equation}

The equilibria are solutions of $f(\hat{p})=0$, or $\hat{p}' = \hat{p}$.
It is easy to verify that fixation of either phenotype, $\hat{p} =  0$ and $\hat{p} = 1$, is an equilibrium.
Since the derivative of $f'(\hat{p}$ is 
\begin{equation}
f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big],
\end{equation}
then the condition for local stability of $\hat{p}=1$ is
\begin{equation} \label{eq:derivative_of_phattag-phat}
	f'(1) =
% 2\cdot\hat{p} (-\alpha bvT + cv(1-T)) + (\alpha bvT -cv(1-T)).
%	(1- 2\hat{p}) (\alpha bvT - cv(1-T)) \mid_{\hat{p}=1} = 
	-\alpha bvT + cv(1-T) < 0,
\end{equation}
which produces the following result.
